---
title: "Correlation Analysis and Dendrograms of Evolutionary and Least Cost Distances for Espeletiinae Individuals in the Eastern Andes"
author: "Sebastián Sánchez Cuartas"
output: 
  html_document:
    toc: true
    toc_depth: 2
    toc_float:
      collapsed: false
      smooth_scroll: true
      position: "left"
    theme: flatly
    highlight: tango
---

------------------------------------------------------------------------

## **Introduction**

This code is designed to generate two correlograms with dendrograms, using evolutionary distances and least-cost distances between individuals of the *Espeletiinae* genus. The goal is to visualize the relationships between individuals based on their evolutionary divergence and cost-distance metrics, and to explore the statistical correlations between these two types of distances. The code produces **two distinct plots**:

1.  A correlogram that shows **Spearman correlation coefficients** between the pairwise distances.

2.  A correlogram representing the **p-values** for the significance of those correlations.

Below, I explain each step in the code and how it contributes to generating these two visualizations.

------------------------------------------------------------------------

## **Reading Distance Matrices**

First, we load two distance matrices:

1.  **Evolutionary Distance Matrix (`evol_dist`)**: This matrix represents the evolutionary distance between individuals, derived from phylogenetic data.

2.  **Least-Cost Distance Matrix (`DistM`)**: This matrix contains the least-cost path distances between individuals, calculated using geographic and environmental variables.

Both matrices are read into R using `read.csv()` and converted into matrix format with `as.matrix()`. Each matrix uses species names as row and column identifiers.

------------------------------------------------------------------------

## **Correlation and Hypothesis Testing**

The next step involves computing pairwise **Spearman correlations** between the evolutionary and least-cost distance matrices. To achieve this, we define the `correlation_test()` function, which:

-   Takes as input the distance values between two individuals from both matrices.

-   Filters out missing or infinite values.

-   Performs a Spearman correlation test (`cor.test()`) to assess the relationship between the two distance types for each pair of individuals.

-   Returns the correlation coefficient and the corresponding p-value.

The results are stored in two matrices: one for the **correlation coefficients** (`cor_matrix`) and another for the **p-values** (`p_values`).

------------------------------------------------------------------------

```{r, echo=TRUE, results='hide', message=FALSE, warning=FALSE}

library(ecodist)
library(gplots)

```

```{r setup, echo=TRUE, results='hide', warning=FALSE}

# Read the distance matrices
evol_dist <- as.matrix(read.csv("C:/Users/Usuario/Desktop/Archivos_tesis/Distancias_evolutivas/Dis_evol_Clado_Colombiano/Dist_evol_Clado_Colombiano_ordenado_COriental.csv", header = TRUE, row.names = 1))
DistM <- as.matrix(read.csv("C:/Users/Usuario/Desktop/Archivos_tesis/Resultados_matrices_indices/Indices_matrices_Clado_Colombiano/DistM.csv", header = TRUE, row.names = 1))

# Number of individuals
n_individuals <- nrow(evol_dist)

# Create matrices to store results
cor_matrix <- matrix(NA, nrow = n_individuals, ncol = n_individuals)
p_values <- matrix(NA, nrow = n_individuals, ncol = n_individuals)

# Define the function to calculate pairwise correlation and perform hypothesis testing
correlation_test <- function(ind1, ind2, dist_matrix1, dist_matrix2) {
  dist1 <- as.numeric(dist_matrix1[ind1, ])
  dist2 <- as.numeric(dist_matrix2[ind2, ])
  
  # Filter NA and Inf
  valid_indices <- complete.cases(dist1, dist2)
  dist1 <- dist1[valid_indices]
  dist2 <- dist2[valid_indices]
  
  # Use cor.test with exact = FALSE to get an approximate p-value in case of ties
  result <- cor.test(dist1, dist2, method = "spearman", exact = FALSE)
  
  return(list(correlation = result$estimate, p_value = result$p.value))
}

# Fill the correlation and p-value matrices
for (i in 1:n_individuals) {
  for (j in 1:n_individuals) {
    if (i != j) {
      # Calculate correlation
      result <- correlation_test(i, j, evol_dist, DistM)
      cor_matrix[i, j] <- result$correlation
      p_values[i, j] <- result$p_value
    }
  }
}

# Assign species names to rows and columns
species_names <- rownames(evol_dist)
rownames(cor_matrix) <- species_names
colnames(cor_matrix) <- species_names

species_names <- rownames(DistM)
rownames(p_values) <- species_names
colnames(p_values) <- species_names

```

------------------------------------------------------------------------

## **Correlogram with Dendrogram (Correlation Coefficients)**

To visualize the relationships, we generate a heatmap with dendrograms using the `heatmap.2()` function. The heatmap displays the pairwise **Spearman correlations** between individuals, with rows and columns representing species names. The dendrograms help cluster individuals based on similarity in their correlation profiles.

The following adjustments are made to the plot:

-   A color gradient from blue to red to represent correlation strength.

-   A dendrogram on both rows and columns to visualize clustering.

-   Proper margins to accommodate the species names.

------------------------------------------------------------------------

```{r heatmap_p_values, echo=TRUE, results='markup', message=FALSE, warning=FALSE}

par(cex.main = 0.8) # Adjust title size

# Create the heatmap with dendrogram using correlations
heatmap.2(as.matrix(cor_matrix),  # Use correlation values as data
          trace = "none",        # No additional color lines
          dendrogram = "both",   # Show dendrogram on both rows and columns
          col = colorRampPalette(c("blue", "white", "red"))(100),  # Change colors if desired
          main = "Spearman Correlation Between Pairs",  # Main title
          key = TRUE,            # Show color key
          density.info = "none", # No density information
          symkey = TRUE,         # Symmetric key in the legend
          labRow = species_names,  # Row labels
          labCol = species_names,  # Column labels
          margins = c(8.5, 8.5))  # Adjust margins to include labels
```

------------------------------------------------------------------------

## **Correlogram with Dendrogram (P-Values)**

A second heatmap is generated to represent the **statistical significance** of the correlations, using **p-values**. This helps assess the reliability of the observed correlations. In this plot, the color gradient moves from blue to green, representing varying levels of p-values, with clustering also visualized through the dendrograms.

------------------------------------------------------------------------

```{r plot, echo=TRUE, results='markup', message=FALSE, warning=FALSE}

par(cex.main = 0.8) # Adjust title size

# Create the heatmap with dendrogram using p-values
heatmap.2(as.matrix(p_values),  # Use p-values as data
          trace = "none",       # No additional color lines
          dendrogram = "both",  # Show dendrogram on both rows and columns
          col = colorRampPalette(c("blue", "white", "lightgreen"))(100),
          main = "P-Values for Spearman Correlation",  # Main title
          key = TRUE,           # Show color key
          density.info = "none", # No density information
          symkey = FALSE,       # No symmetry in the legend
          labRow = species_names,  # Row labels
          labCol = species_names,  # Column labels
          margins = c(8.5, 8.5))   # Adjust margins to include labels

```
